---
title: "Integrating corVis with other packages"
author: "Amit Chinwan and Catherine Hurley"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
     toc: true
vignette: >
  %\VignetteIndexEntry{Integrating corVis with other packages}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r global options, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE,echo=FALSE,fig.height = 5, fig.width = 5,
                      fig.align = 'left')
```

This vignette provides information on how data structures or display methods in _corVis_ can be used with other R packages. We separate out the calculation and visualisation steps in _corVis_, which makes it an open ended package for both data structures and displays.

```{r, setup}
if (!requireNamespace("palmerpenguins", quietly = TRUE))
  install.packages("palmerpenguins")

library(corVis)
library(palmerpenguins)
library(dplyr)
library(kableExtra)
library(ggplot2)
```

## Using data structures from `corVis` with other packages

The package provides a collection of various measures of association which can be used to quantify the relationship between two variables and could be used to explore patterns prior to modeling. The measures available in the package are not limited to *numeric* variables only and can be used with *categorical* and *ordinal* variables as well. Table 1 lists the different measures of association provided in the package with the variable types they can be used with, the package used for calculation, the information on whether the measure is symmetric, and the minimum and maximum value of the measure.


```{r}
measures <- dplyr::tribble(
  ~name, ~nn, ~ff, ~oo, ~nf, ~from, ~range,
  "tbl_cor", "y", " ", " " ," ", "stats::cor", "[-1,1]",
  "tbl_dcor", "y", " ", " ", " ", "energy::dcor2d", "[0,1]",
  "tbl_mine", "y", " ", " ", " ", "minerva::mine", "[0,1]",
  "tbl_ace", "y", "y", " ", "y", "corVis", "[0,1]",
  "tbl_cancor", "y", "y", " ","y", "corVis", "[0,1]",
  "tbl_nmi",  "y", "y", " ", "y", "linkspotter::maxNMI", "[0,1]",
  "tbl_polycor", " ", " ","y", " ", "polycor::polychor", "[-1,1]",
  "tbl_tau", " ", " ","y", " ", "DescTools::KendalTauA,B,C,W", "[-1,1]",
  "tbl_gkGamma", " ", " ", "y", " ", "DescTools::GoodmanKruskalGamma", "[-1,1]",
  "tbl_gkTau", " ", " ", "y", " ", "DescTools::GoodmanKruskalTau", "[0,1]",
  "tbl_uncertainty", " ",  "y", " ", " ", "DescTools::UncertCoef", "[0,1]",
  "tbl_chi", " ",  "y", " ", " ", "DescTools::ContCoef", "[0,1]"
)

kableExtra::kbl(measures,booktabs = T) %>% kableExtra::kable_styling(latex_options = "scale_down")
```
Table 1: Table of association measures available in the package corVis

An analyst can follow *Calculating associations using corVis* vignette for calculating association measures. These data structures are then used with other packages to display association measures.

### Visualising association measures using `corrplot`

The package _corrplot_ provides an implementation of the methods in Friendly (2002) and produces correlation display in matrix layout. The correlation displays in the package are only applicable to numeric variables in a dataset. In _corVis_, we calculate association measures for all variables irrespective of their type and then display these measures with different layouts. After calculating measures we can use other packages which provide correlation displays and support tidy structure of association measures to visualize association in a dataset. Figure 1 shows the correlation plot of the numeric variables in penguins data on the left and association measure display on the right. The plot on the right show strong association for species and other numeric variables which is not present in the left-hand plot.

```{r , corrplot_assoc,echo=TRUE, fig.show="hold", out.width="40%", fig.align='default'}
library(corrplot)
df <- penguins
df_num <- df |>
  dplyr::select(where(is.numeric))
  
corrplot(cor(df_num,use = "pairwise.complete.obs"))

penguins_assoc <- calc_assoc(df)

penguins_assoc_mat <- matrix_assoc(penguins_assoc)

corrplot(penguins_assoc_mat, diag = F)
```

Figure 1: Left: Correlation plot for penguins data using `corrplot`. Right: Association plot for penguins data using association measures calculated by `corVis` and visulaized with `corrplot`.

### Visualising association measures using `linkspotter`

The `linkspotter` package calculates and visualizes association for numeric and factor variables using a network layout plot. The nodes show the variables and the edges represent the measure of association between pair of variables. The width of edge maps to absolute correlation among the nodes. Two nodes are connected if the `corMethod` is defined for the variable pairs. 

It is not possible to use association measures calculated using `corVis` to display with `linkspotter` package as they have hard-coded the displays with measures available only in their package. We create a custom function to display association measures in a network display as shown in Figure 2.  
```{r, linkspotter, echo=TRUE}
library(linkspotter)
library(ggraph)
library(igraph)

corCouples<-multiBivariateCorrelation(df)
linkspotterGraph(corDF = corCouples, corMethod = "pearson", smoothEdges = FALSE, 
  dynamicNodes = FALSE)

nw_corVis <- function(assoc){
  assoc <- sym_assoc(assoc)
  graph <- graph_from_data_frame(assoc)
  ggraph(graph = graph) + 
    geom_edge_link(aes(edge_alpha=abs(measure),color=measure,edge_width=abs(measure))) +
    geom_node_point(size=3) + 
    geom_node_text(aes(label=name),repel = TRUE) +
    guides(edge_alpha="none",edge_width="none") +
    scale_edge_color_gradient2(limits=c(-1,1),low="blue",mid = "white",high="brown")
  
}

nw_corVis(penguins_assoc)
```

Figure 2: Left:Association display for penguins data using `linkspotter`. Right: Association display for penguins data in a network plot.

### Visualising association using `corrr`

Another package useful for exploring correlations is `corrr`. The main focus of the package is to create dataframes of correlation instead of matrices, which then can be explored by leveraging tools from *tidyverse*. 
Similar to `corrplot`, the packge `corrr` doesn't support variables other than numeric. An analyst can calculate association measures for every variable pair using `corVis` and then display these measures using the visualizations offered by the `corrr`.

```{r, corrr,echo=TRUE}
library(corrr)
cor_df <- correlate(df_num)
glimpse(cor_df)
rplot(cor_df) + 
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle=30, hjust=1))

penguins_assoc_df <- as.data.frame(penguins_assoc_mat)
rplot(penguins_assoc_df) + 
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle=30, hjust=1))
```

Figure 3: Left: Association display for penguins data using `corrr`. Right: Assocaition display for penguins data using measures from `corVis` and displayed using `corrr`


## Using displays from corVis with data structures from other packages

### Data structure using `linkspotter`

As mentioned earlier, `linkspotter` package calculates association measures for numeric and factor variables. The measures available in the package are: `pearson`, `spearman`, `kendall`, `mic` and `maxNMI`. We can convert their data structure into _pairwise_ or _multi_pairwise_  depending on the number of measures calculated for every variable pair.

The code snippet below show how we can convert the data structure from `linkspotter` to _pairwise_ and then use plotting function for displaying the association measure.
```{r linkspotter_calc_assoc}
corCouples<-multiBivariateCorrelation(df, corMethods = "pearson")

as_pairwise <- function(corCouples){
  
  pick <- stats::complete.cases(corCouples)
  d <- corCouples[pick,]
  d <- d |>
    select(X1,X2,pearson)
  names(d) <- c("x","y","measure")
  d$measure_type <- "pearson"
  row.names(d) <- NULL
  
  class(d) <- append("pairwise", class(d))
  
  d
}

penguins_assoc <- as_pairwise(corCouples)
plot_assoc_matrix(penguins_assoc)
```
Figure 4: Association plot with measures calculated using `linkspotter`

The below code snippet shows how we can convert a data structure with multiple measures calcultaed using `linkspotter` to _multi_pairwise_ data structure to be used with plotting functions in `corVis` package.
```{r linkspotter_calc_assoc_multi}
corCouples<-multiBivariateCorrelation(df)

as_multi_pairwise <- function(corCouples){
  
  corCouples <- tidyr::pivot_longer(corCouples,cols = 5:9,
                                    names_to = "measure_type",values_to = "measure") |>
    dplyr::select(X1,X2,measure_type, measure)
  pick <- stats::complete.cases(corCouples)
  d <- corCouples[pick,]
  
  names(d)[1:2] <- c("x","y")
  row.names(d) <- NULL
  
  class(d) <- append(c("multi_pairwise", "pairwise"), class(d))
  
  d
}

penguins_assoc_multi <- as_multi_pairwise(corCouples)
plot_assoc_matrix(penguins_assoc_multi)


```

Figure 5: Multiple association measures plot with measures calculated using `linkspotter`
