---
title: "Association measures for mixed variables"
author: "A. Chinwan, C. Hurley"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Association measures for mixed variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup}
library(corVis)
library(mlr3data)
library(MLDataR)
library(palmerpenguins)
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,fig.align = "center",fig.height = 10,fig.width = 10)
```

Correlation displays are popular tools for exploring a multivariate dataset but are only limited to numeric variables. This package extends these displays to nominal and mixed variables and also proposes new visualisation methods for exploring the relationship between all the pairs of variables. We used *Normalised Mutual Information (nmi)* and *Cramer's V* as the measures of association to explore interesting bivarite patterns.

Mutual Information is calculated between two random variables and measures the reduction in uncertainity for one variable given the other variable. It is a measure of dependence between two variables and is symmetrical. The *Normalized Mutual Information* measure is a normalised variant of mutual information and takes up value in the interval $[0,1]$. As it is a symmetric measure, it can be compared with other association measures to discover interesting patterns among the variables.

*Cramer's V* is used to measure association between two nominal variables and has a value in the interval $[0,1]$. It is calculated using the Chi-squared statistic, sample size and minimum number of levels of the two nominal variables. 


## Association displays


```{r}
df <- penguins
df <- as.data.frame(df)
df <- df[complete.cases(df),]

nmi_measure <- association_measure(df, method = "nmi", group_by = "species")

bar_display(nmi_measure)
```


```{r,eval=FALSE}
fact_vars <- names(df)[sapply(df,is.factor)]

cramers.v_measure <- association_measure(df[fact_vars], method = "cramers.v", group_by = "species")

bar_display(cramers.v_measure)
```










