---
title: "corVis: An R Package for Visualising Associations and Conditional Associations"
abstract: >
  Correlation matrix displays are important tools to explore multivariate datasets prior to modeling. These displays with other measures of association can summarize interesting patterns to an analyst and assist them in framing right questions while performing exploratory data analysis. In this paper, we present new visualisation techniques to visualise association between all the variable pairs in a dataset in a single plot, which is something existing displays lack. We extend these displays to regression and classification settings, where these could be used to find out variables with high predictive power. Also, we propse new methods to visualise trivariate relationship summaries using conditioning. We use different layouts like: matrix or linear, to name a few, for our displays which have their own advantages and disadvantages. We use seriation in our displays which helps in highlighting interesting patterns easily. The R package \emph{corVis} provides an implementation.
draft: true
author:  
  
  - name: Amit Chinwan
    affiliation: Maynooth University
    address:
    - Hamilton Institute
    - Maynooth, Ireland
    email:  amit.chinwan.2019@mumail.ie
  - name: Catherine Hurley
    email: catherine.hurley@mu.ie
    affiliation: Maynooth University
    address:
    - Department of Mathematics and Statistics
    - Maynooth, Ireland
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(corVis)
library(palmerpenguins)
library(ggplot2)
library(plotly)
```

# Introduction

Finding association among variables in a multivariate analysis is an important step and can help an analyst to understand and frame right questions about the data. In order to explore these relationships, the analyst could use visual tools like correlation matrix plots (popularly known as corrgrams [@friendly2002corrgrams]) to find out associated variables. These displays are generally used with Pearson's correlation coefficient and are therefore limited to only numeric features. In this paper, we present new visualization techniques for displaying association which will include different variable types, can compare multiple measures of association and can conditionally visualize association at different levels of a factor variable.

There have been extensions to corrgrams like: [@buja2016visualization] and [@sCorrPlot], which have been proposed mainly for exploring correlations among the numeric variables for a high dimensional dataset. We introduce a display which includes all the variables of a dataset, irrespective of the data type, in a conventional corrgram plot displaying  every pairwise association. This saves the effort and time of an analyst for exploring relationship among all the variable pairs. [@kuhn2013applied] have proposed display techniques to compare multiple association measures for every pair of output variable and a predictor to measure the importance of each predictor. This can help in summarizing a complex relationship more efficiently as compared to using just one measure like Pearson's correlation which can only find linear associations. In a similar way, we propose different visualization techniques to compare multiple association measures for all the variable pairs in a dataset which can assist a user in finding interesting patterns.








# Measures of Association

An association measure can be defined as a numerical summary quantifying relationship between two or more variables. For example, Pearson's correlation coefficient summarizes the strength and direction of the linear relationship present between two \emph{numeric} variables and is in the range $[-1,1]$. Similarly, distance correlation coefficient measures the non-linear association between two \emph{numeric} variables and summarizes it in $[0,1]$ where $0$ suggests no non-linear relationship and $1$ suggests very high non-linear relationship. The package provides a collection of various measures of association which can be used to quantify the relationship between two variables and could be used to explore patterns prior to modeling. The measures available in the package are not limited to \emph{numeric} variables only and can be used with \emph{categorical} and \emph{ordinal} variables as well.
\begin{itemize}
\item Pearson's correlation : Describe pearson's correlation.
\item Spearman's rank correlation.
\item Kendall's rank correlation.
\item Distance correlation.
\item Canonical correlations.
\item Maximal-information based non-parametric exploration (MINE) statistics.
\end{itemize}

# Visualising Association

Correlation matrix plots (also called corrgrams) are popular tools for exploring bivariate associations in multivariate datasets prior to modelling. These displays are generally used with Pearson’s correlation coefficient and are thus suitable for numeric variables only. We propose novel visualisations to display association for every variable pair in a dataset in a single plot and show multiple bivariate measures of association simultaneously to find out interesting patterns.

We propose a display for visualising association for different variable types. This is an extension of existing corrgram which is only suitable for numeric variables in a dataset. We calculate different assocaition measures for different types of variable pairs and then plot them in a similar way as it is done in a conventional corrgram. The traditional method to find interesting bivariate relationships is to split the dataset by the variable types and then analyse these one by one. Our approach saves the effort and time of an analyst for exploring pairwise relationships among all the variables in the dataset and can be done in a single step.

We introduce a new structure for calculating association measures which can be used to add other existing or new measures in the package. These measures can then then be analysed and visualised using the plot functions present in the package. For example, Cramer's V is a measure to summarize association between two categorical variables using the Chi-square test statistic. If a user wants to add Cramer's V to the package, they can write a simple function and then can use it for their analysis.

We consider matrix-type, linear and network-based layouts. A matrix-type layout simplifies lookup, and different measures may be displayed on the upper and lower diagonal. Linear layouts are more space-efficient than matrix plots, but lookup is more challenging. Variable pairs can be ordered by relevance (usually difference in measures of association or across the factor levels), and less relevant pairs can be omitted. Linear displays are also suitable to display associations between the response and predictors only. Our selection criteria for a better display were based on :
\begin{itemize}
\item Number of variables
\item Easier pixel-variable or variable-pixel
look up
\item Number of levels of a factor for conditional association displays
\end{itemize}

Figure \@ref(fig:assoc-heatmap) shows this display for every variable pair in the *penguins* dataset from the *palmerpenguins* package. It shows a high positive Pearson's correlation among flipper_length_mm and body_mass_g, flipper_length_mm and bill_length_mm, and bill_length_mm and bodymass_g. There seems to be a strong negative Pearson's correlation between flipper_length_mm and bill_depth_mm, and bill_depth_mm and body_mass_g. The plot also shows that there is a high canonical correlation between species and other variables except year and sex, and a high canonical correlation between island and species, which traditional correlation matrix display would omit as they are limited to numeric variable pairs only. The variables in the display are ordered using average linkage clustering method to find out highly associated variables quickly.

```{r assoc-heatmap,fig.width=6, fig.height=6, fig.align='center', fig.cap="Association matrix display for penguins data"}
df <- penguins
assoc <- calc_assoc(df)
association_heatmap(assoc)
```


We can also calculate multiple association measures for all the variable pairs in the dataset and compare them. This will help in finding out pairs of variables with a high difference among different measures and one can investigate these bivariate relationships in more detail. The `pairwise_summary_plot` function can be used to compare various measures using the matrix layout. It plots multiple measures among the variable pairs as bars, where each bar represents one measure of association. Figure \@ref(fig:compare-matrix) shows a matrix layout comparing Pearson's and Spearman's correlation coefficient for the numeric variable pairs in *penguins* data. 

```{r compare-matrix, fig.width=6, fig.height=6, fig.align='center',fig.cap="Comparing Pearson's and Spearman's correlation coefficient"}

df_num <- select(df,where(is.numeric))
pearson <- calc_assoc(df_num)

spearman_assoc <- update_assoc(num_pair = "tbl_cor",
                               num_pair_argList= "spearman",
                               mixed_pair = "tbl_cancor",
                               other_pair = "tbl_nmi")
spearman <- calc_assoc(df_num, types=spearman_assoc)
compare <- rbind(pearson,spearman)
pairwise_summary_plot(compare, group_var = "measure_type")
```

In addition to matrix layout, we can also use linear layouts for comparing multiple measures. Figure \@ref(fig:compare-linear) shows a linear layout comparing multiple association measures for all the variable pairs in the penguins data.  Linear layouts seems to be more suitable when comparing high number of association measures.   

```{r compare-linear,fig.width=6, fig.height=6, fig.align='center',fig.cap="Comparing multiple association measures using a linear layout"}


pairwise_measures_compare(df,measures=c("pearson","spearman","kendall","cancor","nmi"))

```



# Visualising Conditional Association

The package includes a function `calc_assoc_by` which calculates the pairwise association at different levels of a categorical conditioning variable. This helps in finding out interesting variable triples which can be explored further prior to modeling. Figure \@ref(fig:cond-assoc) shows a conditional association plot for the *penguins* data. Each cell corresponding to a variable pair shows three bars which correspond to the association measure (Pearson's correlation for numeric pair and Normalized mutual information for other combination of variables) calculated at the levels of conditioning variable *island*. The dashed line represents the overall association measure. The plot shows that there is a high value for normalised mutual information between bill_length_mm and species for the penguins which lived in *Biscoe* island compared to the penguins which lived in *Dream* island. It can also be seen that the cell corresponding to variable pair flipper_length_mm and bill_depth_mm has a high negative overall Pearson's correlation and for the penguins which lived in *Biscoe* island but positive  correlation for penguins which lived in *Dream* and *Torgersen* island. This is an instance of Simpson's paradox which can be taken into account during the modeling step.


```{r cond-assoc,fig.width=6, fig.height=6, fig.align='center', fig.cap="Figure 4: Conditional Association plot"}
updated_assoc <- update_assoc(num_pair = "tbl_cor",
                              mixed_pair = "tbl_nmi",
                              other_pair = "tbl_nmi")

cond_assoc <- calc_assoc_by(df, by="island")
pairwise_summary_plot(cond_assoc)
```

# Discussion

We have displayed various tooltips that are available in the package \pkg{ToOoOlTiPs}. 
