---
title: "corVis: An R Package for Visualising Associations and Conditional Associations"
abstract: >
  Correlation matrix displays are important tools to explore multivariate datasets. These displays with other measures of association can summarize interesting patterns to an analyst and assist them in framing questions while performing exploratory data analysis. In this paper, we present new visualisation techniques to visualise association between all the variable pairs in a dataset in a single plot, which is something existing displays lack. Also, we propse new methods to visualise relationship among variable pairs using conditioning. We use different layouts like matrix or linear for our displays. We use seriation in our displays which helps in highlighting interesting patterns easily. The R package `corVis` provides an implementation.
draft: true
author:  
  
  - name: Amit Chinwan
    affiliation: Maynooth University
    address:
    - Hamilton Institute
    - Maynooth, Ireland
    email:  amit.chinwan.2019@mumail.ie
  - name: Catherine Hurley
    email: catherine.hurley@mu.ie
    affiliation: Maynooth University
    address:
    - Department of Mathematics and Statistics
    - Maynooth, Ireland
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(corVis)
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(kableExtra)
```

# Section 1: Introduction

Correlation matrix display is a popular tool to visually explore correlations among variables while performing Exploratory Data Analysis (EDA) on a multivariate dataset. Popularized by @friendly2002corrgrams as corrgram, these displays are produced by first calculating the correlation among the variables and then plotting these calculated values in a matrix display. With effective ordering techniques, these displays quickly highlight variables which are highly correlated and an analyst interested in building a predictive model could use these displays to remove correlated variables and avoid multicollinearity.

The correlation displays are generally used with one of the Pearson's, Spearman's or Kendall's correlation coefficient and are therefore limited to quantitative variables. An analyst can use one-hot encoding of the qualitative variables in order to use these displays but will need to deal with the high dimensions as a result of the encoding. In addition to the dimensionality problem, it is not easy to assess the overall correlation when using the one-hot encoding.  The existing methods to quickly explore association among qualitative variables in a dataset include using proportions or counts with different graphical displays like boxplots or barplots. Using association measures for qualitative pairs similar to correlation for quantitative pairs will help in summarizing the relationship, which then can be displayed like the correlation displays. 

Tukey and Tukey introduced scagnostics which are measures for scatterplots [@tukey1985computer]. Along with scagnostics, they proposed a scagnostics scatterplot matrix which is a visual display to explore and compare these measures for all the variable pairs in a dataset. By comparing multiple measures at once, the unusual variable pairs could be identified and looked at in more detail. In a similar manner, a display comparing association measures will help in finding interesting variable pairs. Many association measures have been proposed to summarize different types of relationships. The most commonly used measure is Pearson's correlation coefficient which captures any linear trend present between the variables. Other popular measures include Kendall's or Spearman's rank correlation coefficient which are non-parametric measures and looks for monotonic relationship. Distance correlation [@szekely2007measuring] is an important measure useful in exploring non-linear relationships. The information theory measure maximal information coefficient (MIC) [@reshef2011detecting] is capable of summarizing complex relationships. With effective displaying techniques, the multiple measures of association provide a comparison tool that assist an analyst to reveal structure present in the data.

Small multiples (or Trellis display) is a simple yet powerful approach to compare partitions of data and understand multidimensional datasets [@tufte1986thevisual]. The display is produced by splitting the data into groups by a conditioning variable and then plotting the data for each group. Such displays allow analysts to quickly infer about the impact of the conditioning variable. A similar idea applied to displays of association measures (correlation plot) will help uncover underlying patterns in the data. One such pattern is Simpson's paradox which can be detected by comparing Pearson's correlation for data at overall level versus individual levels of the conditioning variable.

In this paper, we propose extensions of the correlation plot and new visualizations which look at variables of mixed type, multiple association measures and conditional associations. These displays are implemented in the R package \CRANpkg{corVis}. The next section provides a review of existing packages which deal with correlation displays and a quick background on association measures and the packages used for calculating them. Then we describe our approach to calculate the association measures, followed by visualizations of associations and conditional associations. We conclude with a summary and future work.

# Section 2: Background

In this section we provide a brief review of  existing packages used for correlation displays and  association measure calculation. 

## Section 2.1: Literature Review on Correlation Displays

<!-- HISTORICAL REFS ADD HERE -->

According to @hills1969looking, "the first and sometimes only impression gained by looking at a large correlation matrix is its largeness". To overcome this, @murdoch1996graphical proposed a  display for  large correlation matrices which uses a matrix layout of ellipses where the parameters of the ellipses  are  scaled to the correlation values. @friendly2002corrgrams  expanded on this idea by rendering correlation values as shaded squares, bars, ellipses, or circular ‘pac-man’ symbols. The variables in the matrix displays were optionally ordered using the angular ordering of the first two eigen vectors of the correlation matrix. The ordering places highly-correlated pairs of variables nearby, making it easier to quickly identify groups of variables with high mutual correlation.

<!-- In order to explore a large correlation matrix @hills1969looking proposed a QQ plot of the entries of the correlation matrix. According to @murdoch1996graphical, they proposed a more effective display for exploring a large correlation matrix. @murdoch1996graphical replaced the entries in a correlation matrix by an ellipse where the parameters of ellipse were scaled to the correlation value. These displays are called as correlation matrix displays. -->

<!-- The correlation matrix display is an important tool to explore association among variables in a multivariate analysis. The display was made popular by @friendly2002corrgrams who called them corrgrams, wherein he rendered the correlation values of $p$ numeric variables in a $p \times p$ matrix layout with shaded squares, bars, ellipses, or circular ‘pac-man’ symbols. The main goal of these displays is to render the correlation patterns in a dataset.  -->

Nowadays, there are many R  packages devoted to correlation visualisation. Table \@ref(tab:corrdisplay-packages) provides a summary, listing the displays offered, and whether these extend to factor variables or mixed numeric-factor pairs. 

The R package \CRANpkg{corrplot} [@corrplot2021] provides an implementation of the methods in @friendly2002corrgrams. 
 The package \CRANpkg{corrr} [@corrr2020] organises correlations as tidy data, so leveraging the data manipulation and visualisation tools of the \CRANpkg{tidyverse} [@tidyverse]. In addition to various matrix displays, the package offers network displays where line-thickness encodes correlation magnitude, with a filtering option to discard low-correlation edges.

<!-- Table \@ref(tab:corrdisplay-packages) provides a list of  packages available in R which  either calculate correlations, visualise correlations or both.  -->
<!-- The displays provided by each of the packages are listed, as are the correlations or associations calculated, in particular whether association measures are provided for factor variables or mixed numeric-factor pairs. We also summarise whether packages provide conditional displays of association, by which we mean displays for each of the levels of a categorical variable. -->

<!-- The R package \CRANpkg{corrplot} [@corrplot2021] provides an implementation of the @friendly2002corrgrams paper. It serves as a visual exploratory tool for correlation matrices and includes various variable ordering methods which place highly-correlated pairs of variables nearby, making it easier to quickly identify groups of variables with high mutual correlation. -->


<!-- The package \CRANpkg{corrr} [@corrr2020] organises correlations as tidy data, so leveraging the data manipulation and visualisation tools of the \CRANpkg{tidyverse} [@tidyverse]. In addition to various matrix displays, the package offers network displays where line-thickness encodes correlation magnitude, with a filtering option to discard low-correlation edges. -->


<!-- CORRGRAPHER -->

The package \CRANpkg{corrgrapher} [@corrgrapher] uses a network plot for exploring correlations, where the nodes close to each other have high correlation magnitude, edge thickness encodes the absolute correlation value and edge color indicates the sign of correlation. The package also handles mixed type variables by using association measures obtained as transformations of $p$-values obtained from
Pearson's correlation test in the case of two numeric variables, Kruskal's test for numerical and factor variables, and a chi-squared test for two categorical variables. 

The package \CRANpkg{linkspotter} [@linkspotter] offers a variety of association measures (distance correlation, MIC, maximum normalized mutual information) in addition to correlation, where the measure used depends on whether the  variables are both numerical, categorical or mixed. The results are visualized in a network plot, which may be packaged into an interactive shiny application.


Our own package \CRANpkg{corVis} offers a variety of displays, and has new features not available elsewhere, in particular simultaneous display of multiple association measures, and association displays stratified by levels of a grouping variable. This will be described  in the following sections.

<!-- ALL THIS STUFF ABOUT CORVIS SHOULD APPEAR IN THE SECTION ABOUT CORVIS. -->

<!-- It allows the user to select association measure for different variable types and then calculates and visualizes all of the variable pairs in a single plot. This helps an analyst to explore the associations among different variables in a single display. -->



There have been other extensions to correlation displays which are useful when dealing with high dimensional datasets. 
@hills1969looking proposed a QQ plot of the $z$-transform of the entries of the correlation matrix to discover correlation coefficients too large to come from a normal distribution with mean zero. @buja2016visualization proposed Association Navigator which is an interactive visualization tool for large correlation matrices with upto 2000 variables.  The R package \CRANpkg{scorrplot} [@scorr] produces an interactive scatterplot for exploring pairwise correlations in a large dataset by projecting variables as points and encoding the correlations as space between these points. The package provides a functionality to update variable of interest which creates tour of the correlation space between different projections of the data. 

<!-- I DO NOT UNDERSTAND ANY OF THIS ABOUT SCORRPLOT. TRY AGAIN. -->

<!-- REMOVE CORHEAT FROM TABLE -->


The R package \CRANpkg{correlationfunnel} offers a novel display which assists in feature selection in a setting with a single response and many predictor variables. All numeric variables including the response are binned. All (now categorical) variables in the resulting dataset are one-hot encoded and Pearson's correlation calculated with the response categories. The correlations are visualised in a dot-plot display, where predictors are ordered by maximum correlation magnitude. Correlations between one-hot encoded variables are challenging to interpret, especially as the number of levels increase.  In corVis we offer a similar dot-plot display, but showing multiple correlation or association measures, or alternatively measures stratified by a grouping variable.


```{r corrdisplay-packages,warning=FALSE,message=FALSE,echo=FALSE }
library(kableExtra)

Package <- c("corrplot","corrr","corrgrapher","linkspotter",  "correlation","corVis")

Display <- c("heatmap","heatmap/network","network","network","heatmap/network","heatmap/matrix/linear")


MixedVariables <- c(" "," "," ","Yes"," ","Yes")



df <- data.frame(Package=Package,
                 Display=Display,
                 MixedVariables=MixedVariables)

kableExtra::kbl(df, booktabs = T, caption = "List of the R packages dealing with correlation or correlation displays with information on whether the plots display multiple measures, conditional display of measures and mixed variables in a single plot")
#print(table_1)
#df %>%
  #kbl() %>%
  #kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```




## Section 2.2: Literature Review on Association Measures

An association measure can be defined as a numerical summary quantifying the relationship between two or more variables. For example, Pearson's correlation coefficient summarizes the strength and direction of the linear relationship present between two numeric variables and is in the range $[-1,1]$. Kendall's or Spearman's rank correlation coefficient are other popular measures which assess montonic relationship among two numeric variables and are in the range $[-1,1]$. As these measures are limited to linear or monotonic relationships, there's a need to use association measures which are able to capture complex relationships. In addition to association measures for numeric variables, association measures for ordinal, nominal and mixed variable pairs are useful in exploring a multivariate dataset. We now give an overview of available association measures.

For a pair of numeric variables, various measures of association have been proposed in literature. The distance correlation coefficient [@szekely2007measuring] is an association measure which looks for the non-linear association between two numeric variables and summarizes it in $[0,1]$. Similarly, MIC [@reshef2011detecting] is capable of summarizing non-linear as well as periodic relationships between numeric variables and is in range $[0,1]$.

@agresti2010analysis provides an overview of the association measures which are used for exploring association between ordinal variables. Kendall's tau-b [@kendall1945treatment] is an association measure useful in summarizing the relationship between two ordinal variables in the range $[-1,1]$. It is a relatively stable measure than Goodman and Kruskal's gamma with respect to the changes in categories of any variable i.e. if two categories are merged to make a single category. The polychoric correlation [@olsson1979maximum] measures the correlation between two ordinal variables  by assuming two normally distributed latent variables for a contingency table of two ordinal variables and summarizes the association in $[-1,1]$.

<!-- For a pair of nominal variables -->
The association measures for the case of nominal pair of variables should be invariant to the order in which the categories appear. Pearson's contingency coefficient uses the ${\chi}^2$ value from the Pearson's ${\chi}^2$ test for independence and is a useful measure to summarize the association between two nominal variables in $[0,1]$.  Another measure for nominal variable pair is the Uncertainty coefficient [@theil1970estimation] measuring the proportion of uncertainty in one variable which is explained by the other variable.

<!-- For a pair of mixed variables -->

<!-- The canonical correlation is a measure useful in exploring association among mixed variables. The goal of the canonical correlation analysis is to maximize the association between the low-dimensional projections of two sets of variables [@hardle2019applied]. Each of these measures are are consistent with respect to the order of the categories of the nominal variable. -->




# Section 3: Introducing corVis

\CRANpkg{corVis} is an R package which calculates measures of association for every variable pair in a dataset and helps in visualising these associations in different ways. Most of the existing correlation displays are limited to numeric pairs of variables. This package extends these displays to every variable pair.

The package offers new visualisation technique such as display with multiple measures for the association for every variable pair in the dataset. We also introduce conditional association displays which are useful in uncovering conditional structure present in the data and are produced by splitting the data by a partitioning variable, calculating association for the variable pairs at every level of partitioning variable and then plotting them. These displays are efficient for discovering differences among the groups in the data.


Efficient seriation techniques have been included to order and highlight variables with high value for an association measure. These ordered association and conditional association displays are useful in finding interesting patterns, such as Simpson's Paradox. These new displays also help an analyst to quickly discover any unusual variable pair(s) in the dataset.

Table \@ref(tab:function-corVis) provides a list of the functions available in the package which are useful for calculating association measures among variable pairs and visualising these associations using novel displays. Section 4 provides a detailed description on functions used for calcuating association measures in the package. Section 5 illustrates the use of visualising functions in table \@ref(tab:function-corVis) for displaying pairwise association and conditional association.


```{r function-corVis,warning=FALSE,message=FALSE,echo=FALSE }
library(kableExtra)

Function <- c("calc_assoc","association_heatmap","pairwise_2d_plot","pairwise_1d_compare","pairwise_1d_plot")

Usage <- c("Calculation","Visualization","Visualization","Visualization","Visualization")

#Usage <- c(rep("Calculation",5),rep("Visualization",3),rep("Visualization",5),rep("Visualization",4),rep("Visualization",5))

Input <- c("Calculates association and conditional association",
           
           "Plots association for every variable pair in a matrix layout",
           
           "Plots conditional association for every variable pair in a matrix layout",
           
           "Plots multiple association measures for every variable pair in linear layout",
           
           "Plots conditional association dotplot for every variable pair in a linear layout")


df <- data.frame(Function=Function,
                 Usage=Usage,
                 Input=Input)

kableExtra::kbl(df, booktabs = T, caption = "List of the available functions in corVis package") %>%
  kableExtra::column_spec(1, bold=T)


  #kableExtra::collapse_rows(columns = 1:2)

#print(table_1)
#df %>%
  #kbl() %>%
  #kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```


# Section 4: corVis: Calculating Association

This section describes the calculation of association measures in our package \CRANpkg{corVis}. The package  provides a standard interface for calculating a collection of various measures of association which quantifies the relationship between two variables. The association measures available in the package are not limited to numeric variables and are used with nominal, ordinal and mixed variable pairs as well. The package also provides a functionality for handling missing value or `NA` while calculating the association measures. 

Table \@ref(tab:association-measures) lists different functions provided in the package to calculate varoius measures of association. The `Function` column represents the function name used to calculate measure(s) of associations in this package. The `typeX` and `typeY` columns provide the information on types of variables which can be used with the corresponding functions. The `X` or `Y` variable is one of the numeric, nominal, ordinal or any type. The `from` column corresponds to the package functions used to calculate the association measures by the function under `Function`. The `symmetric` column represents if the measure is symmetric i.e. if the value of measure is same regardless of the order of variables. The last column provides the range of values for these measures. The function `tbl_easy` calculates association measures available in the R package \CRANpkg{correlation} and is suitable for different variable types. The functions in Table \@ref(tab:association-measures) with `corVis` entries under `from` column calculate the association measures which have been implemented in this package. 


For numeric pairs of variables, this package provides a range of association measures. The popular correlation coefficients like Pearson's or Spearman's or Kendall's are calculated using `tbl_cor` function. The measures such as distance correlation or MIC are calculated using `tbl_dcor` or `tbl_mine` respectively. The association measures  available in the package for the ordinal pairs of variables are polychoric correlation and Kendall's coefficients which are calculated using `tbl_polycor` or `tbl_tau` respectively. For nominal pairs of variables, the functions like `tbl_gkTau`, `tbl_gkGamma`, `tbl_uncertainty`, `tbl_chi`, `tbl_cancor` are used for exploring association among the variables. 

<!-- These measures are consistent with respect to the order of the levels of nominal variable which some of the existing measures lack. -->

The function `tbl_cancor` calculates a measure of association based on canonical correlations for mixed pairs of variables. Nominal variables are converted into sets of dummy variables, which are then assigned scored to find the maximal correlation. For two numeric variables this measure is identical to absolute correlation, for two factors the correlation is identical to that obtained from correspondence analysis.

The functions listed in \@ref(tab:association-measures) for calculating association measures provide a functionality for handling missing value or `NA` in the dataset. Each of these functions either have a `handle.na` argument or have package functions which automatically uses pairwise complete observations for taking care of missing values present in the data. 

```{r association-measures,echo=FALSE}
assocMethods <- tribble(
  ~Function, ~X, ~Y,  ~from, ~symmetric, ~range,
  "tbl_cor", "numerical", "numerical",  "stats::cor", "Y", "[-1,1]",
  "tbl_dcor", "numerical", "numerical",  "energy::dcor2d", "Y", "[0,1]",
  "tbl_mine", "numerical", "numerical", "minerva::mine","Y", "[0,1]",
  "tbl_polycor", "ordinal", "ordinal", "polycor::polychor", "Y", "[-1,1]",
  "tbl_tau", "ordinal", "ordinal", "DescTools::KendalTauA,B,C,W", "Y", "[-1,1]",
  "tbl_gkGamma", "ordinal", "ordinal", "DescTools::GoodmanKruskalGamma", "Y", "[0,1]",
  "tbl_gkTau", "nominal", "nominal", "DescTools::GoodmanKruskalTau", "N", "[0,1]",
  #"tbl_gkLambda", "nominal", "nominal", "DescTools::GoodmanKruskalTau", TRUE, "[0,1]",
  "tbl_uncertainty", "nominal", "nominal", "DescTools::UncertCoef", "Y", "[0,1]",
  "tbl_chi", "nominal", "nominal", "DescTools::ContCoef", "Y", "[0,1]",
  "tbl_cancor", "nominal/numerical", "nominal/numerical", "corVis", "Y", "[0,1]",
  "tbl_nmi", "any", "any", "corVis", "Y", "[0,1]",
  "tbl_easy", "any", "any", "correlation::correlation", "Y", "[-1,1]")

kableExtra::kbl(assocMethods,booktabs = T,caption = "List of the functions available in the package for calculating different association measures along with the packages used for calculation.")
#kableExtra::kable_styling(kable_input,latex_options = "striped")
#knitr::kable(assocMethods)
```

## Calculating association for a single type of variable pairs

We have a function which creates a tibble structure for the variable pairs in a dataset along with calculated association measure. The package contains various functions (shown in Table \@ref(tab:association-measures)) for different association measures in the form `tbl_*` to calculate them. For example, in order to calculate distance correlation for numeric pair of variables in a dataset, the function `tbl_dcor` is used.

```{r,echo=TRUE}
ckd <- RWeka::read.arff("../data/chronic_kidney_disease.arff")
df <- ckd
#df$al <- as.ordered(df$al)
#df$su <- as.ordered(df$su)

distance <- tbl_dcor(df)
distance
```


The tibble output for the functions mentioned in Table \@ref(tab:association-measures) has the following structure:

  - `x` and `y` representing a pair of variables
  - `measure` representing the calculated value for association measure
  - `measure_type` representing the association measure calculated for `x` and `y` pair.

## Calculating association measures for whole dataset

`calc_assoc` is used to calculate association measures for all the variable pairs in the dataset at once in a tibble structure. The variable pairs in the output are unique pairs and a subset of all the pairs of variables in a dataset where `x` $\neq$ `y`. Because of the tidy structure of the output, the data manipulation and visualisation tools of \CRANpkg{tidyverse} [@tidyverse] are applicable to and are useful for further exploration of pairwise associations. In addition to tibble structure, the output also has `pairwise` and `data.frame` class which are important class attributes for producing visual summaries in this package.

The function `calc_assoc` has a `types` argument which is basically a tibble of the association measure to be calculated for different variable pairs. The default tibble of measures is `default_assoc()` which calculates Pearson's correlation if both the variables are numeric, Kendall's tau-b if both the variables are ordinal, canonical correlation if one is factor and other is numeric and canonical correlation for the rest of the variable pairs.

```{r,echo=TRUE}
default_measures <- default_assoc()
default_measures

ckd_assoc <- calc_assoc(df,types = default_assoc())
ckd_assoc
class(ckd_assoc)
```


The default tibble of measures is updated using the `update_assoc` function which has arguments for updating the `tbl_*` functions to calculate association measures depending on the type variable pair in the dataset and a method for `tbl_*` functions which calculates more than one measure. The `update_assoc` function has an argument `default` which has the `default_assoc()` tibble as its default value and is useful when `tbl_*` functions need to be updated for a few types of variable pairs. 


```{r,echo=TRUE}
updated_assoc <- update_assoc(default=default_assoc(),
                              num_pair = "tbl_cor",
                              num_pair_argList = "spearman",
                              mixed_pair = "tbl_cancor",
                              other_pair = "tbl_nmi")
updated_assoc

```

`calc_assoc` also has a `handle.na` argument for handling the `NA` or missing values which is fed into the `tbl_*` functions used with the `types` argument for different types of variable pairs. The default value is set to `TRUE` for using pairwise complete observations for calculating a measure of association between two variables.



If a user is interested in calculating multiple association measures for a type of variable pair, it can be done by using the `calc_assoc` and `update_assoc` together for calculating different association measures and then merging the output tibbles.

```{r,echo=TRUE}

updated_ckd_assoc <- calc_assoc(df, types = updated_assoc)
updated_ckd_assoc

```



## Calculating conditional association

`calc_assoc` is also used to calculate association measures for all the variable pairs at different levels of a categorical variable. This helps in exploring the conditional associations and find out the differences between the groups of the conditioning variable. The function has a `by` argument which is used as the grouping variable and needs to be categorical.

```{r,echo=TRUE}
ckd_assoc_by <- calc_assoc_by(df,by = "htn")
ckd_assoc_by
```

<!-- The `calc_assoc_by` function also has a `types` argument which can be updated similarly to `calc_assoc`.  -->

<!-- ```{r,echo=TRUE} -->
<!-- updated_assoc <- update_assoc(num_pair = "tbl_cor", -->
<!--                               num_pair_argList = "spearman", -->
<!--                               mixed_pair = "tbl_cancor", -->
<!--                               other_pair = "tbl_nmi") -->
<!-- updated_penguin_assoc_by <- calc_assoc_by(df,by = "sex", types = updated_assoc) -->
<!-- head(updated_penguin_assoc_by) -->
<!-- ``` -->

By default, the function `calc_assoc` calculates the association measures for all the variable pairs at different levels of the grouping variable and the pairwise association measures for the ungrouped data (`overall`) when used with the `by` argument. This behavior can be changed by setting `include.overall` argument to `FALSE`.

```{r,echo=TRUE}
ckd_assoc_by <- calc_assoc_by(df,by = "htn",include.overall = FALSE)
ckd_assoc_by
```

The tibble output in the conditional setting has a similar structure as `calc_assoc` used with no `by` argument. When used with the `by` argument, an additional `by` column representing the levels of the categorical variable is added in the tibble output. The `x` and `y` variables in the output are repeated for every level of `by` variable. In order to have multiple `by` variables, the function `calc_assoc` is used multiple times with a different `by` variable each time and then the multiple outputs are binded row wise. For calculating multiple measures for a specific variable type, one can use `update_assoc` with `calc_assoc` and then can merge these multiple tibble outputs.



# Section 5: corVis: Visualising Association

This section provides a detailed description of the novel visualisation techniques proposed in the package \CRANpkg{corVis}. These methods display association and conditional association for every variable pair in a dataset in a single plot and show multiple bivariate measures of association simultaneously. 

While designing these displays we considered matrix, linear and network-based layouts. A matrix layout simplifies the effort in finding variables, and different measures may be displayed on the upper and lower diagonal. Linear layouts are more space-efficient than matrix plots, but looking for variables or variable pairs is more challenging. The variable pairs in both matrix and linear layouts are usually ordered by relevance (usually the association measure difference in measures of association or across the factor levels for conditional displays) so that pairs of variables having strong association, with high difference among groups or with any other pattern are esaily higlighted.

We use chronic kidney dataset providing information on early stage of Chronic Kidney Disease(CKD) patients from the [@Dua:2019] to provide illustrative examples. Table \@ref(tab:ckd) provides a brief description of set of variables from this dataset used for analysis.


```{r ckd,warning=FALSE,message=FALSE,echo=FALSE }
library(kableExtra)

Variable <- c("bgr","bu","rbcc","pcv","sod","sc","al","su","htn","dm","cad")

Description <- c(
                 "Blood Glucose Random in mgs/dl",
                 "Blood Urea mgs/dl",
                 "Red Blood Cell Count in millions/cmm",
                 "Packed  Cell Volume",
                 "Sodium in mEq/L",
                 "Serum Creatinine in mgs/dl",
                 "Albumin (0,1,2,3,4,5)",
                 "Sugar (0,1,2,3,4,5)",
                 "Hypertension (yes,no)",
                 "Diabetes Mellitus (yes,no)",
                 "Coronary Artery Disease (yes,no)")

VariableType <- c("numerical","numerical","numerical","numerical","numerical","numerical","ordinal","ordinal","nominal","nominal","nominal")

df <- data.frame(Variable=Variable,
                 Description=Description,
                 VariableType=VariableType)

kableExtra::kbl(df, booktabs = T, caption = "Variable description of the chronic kidney dataset along with the types of variables")
#print(table_1)
#df %>%
  #kbl() %>%
  #kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```


## Association Matrix plot

The function `association_heatmap` is used to display a matrix layout with association for variable pairs in the dataset. The display is similar to existing correlation matrix plots but with every variable pair in the dataset. This function `association_heatmap` takes the calculated measures of association by `calc_assoc` function as input and outputs a matrix display by rendering the magnitude of association measures with a color. The function has `lassoc` and `uassoc` arguments for a tibble of association measures for the lower triangle and the upper triangle of the matrix display respectively. The `uassoc` argument is `NULL` by default and uses the same tibble input as used by `lassoc` if not changed. The argument `var_order` is used for ordering or seriating the matrix display such that highly-associated variables are placed nearby and are easier to identify. We use average linkage hierarchical clustering method as the default method for ordering the variables. The function also has a `limits` argument specifying the limit of the color scale. 



Figure \@ref(fig:assoc-heatmap) shows an example of the association matrix display for every variable pair in the chronic kidney disease dataset. The cells along the diagonal of the matrix display show the variables present in the dataset. Every off diagonal cell is colored using a divergent color scale with limits $[-1,1]$ representing the value of association measure between two variables. The plot shows Pearson's correlation for the numeric pair of variables and canonical correlation for mixed and nominal variable pairs. 


```{r assoc-heatmap,fig.width=5, fig.height=4, fig.align='center', fig.cap="Association matrix display for penguins data showing Pearson's correlation for numeric variable pairs, canonical correlation for mixed variable pairs and categorical variable pairs.",echo=TRUE}
ckd <- RWeka::read.arff("../data/chronic_kidney_disease.arff")

vars <- c("bgr","bu","rbcc","pcv","sod","sc","al","su","htn","dm","cad")
df <- ckd[,vars]
#df$al <- as.ordered(df$al)
#df$su <- as.ordered(df$su)

assoc <- calc_assoc(df)
association_heatmap(assoc)
```


For numeric pairs, Figure \@ref(fig:assoc-heatmap) shows a high positive Pearson's correlation among `rbcc` and `pcv` and a high negative Pearson's correlation between `sc` and `sod`. Figure \@ref(fig:numeric_pairs) shows scatterplot for these two pairs of variables and is evident that these pairs seems to be linearly associated.

```{r numeric_pairs, fig.width=6, fig.height=5, fig.align='center', fig.cap = "Scatterplot for variables rbcc and pcv, and, sc and sod showing a high positive and negative Pearson's correlation respectively."}

p1 <- ggplot2::ggplot(data = df) + ggplot2::geom_point(ggplot2::aes(y=rbcc,x=pcv))

p2 <- ggplot2::ggplot(data = df) + ggplot2::geom_point(ggplot2::aes(y=sc,x=sod))

gridExtra::grid.arrange(p1,p2,ncol=2)

```


Figure \@ref(fig:assoc-heatmap) shows a high canonical correlation between `bgr` and `su`, `rbcc` and `htn`, and, `pcv` and `al` which are mixed variable pairs in the dataset. These variable pairs are further looked into using boxplots. Figure \@ref(fig:mixed_pairs) shows boxplot for these variable pairs and is evident there is an association among these variable pairs.

```{r mixed_pairs, fig.width=6, fig.height=5, fig.align='center', fig.cap = "Boxplot for variables bgr and su, htn and rbcc, and, al and pcv showing a high canonical correlation . "}

p3 <- ggplot2::ggplot(data = df) + ggplot2::geom_boxplot(ggplot2::aes(y=bgr,x=su))

p4 <- ggplot2::ggplot(data = df) + ggplot2::geom_boxplot(ggplot2::aes(y=rbcc,x=htn))

p5 <- ggplot2::ggplot(data = df) + ggplot2::geom_boxplot(ggplot2::aes(y=pcv,x=al))

gridExtra::grid.arrange(p3,p4,p5,ncol=3)
```

For nominal variable pairs, Figure \@ref(fig:assoc-heatmap) shows a high canonical correlation among `htn` and `al`, `htn` and `dm`, and, `dm` and `su`. These pairs of variables can be further looked at by plotting proportional bar charts.




## Multiple Association Measures Plot

<!-- We also calculate multiple association measures for all the variable pairs in the dataset and compare them. This helps in identifying pairs of variables with a high difference among different measures or any unusual variable pairs which should be investigated further in more detail.  -->

The function `pairwise_1d_compare` is used to display a linear layout for comparing multiple association measures calculated for variable pairs in the dataset. `pairwise_1d_compare` takes a dataset and a set of association measures as input and outputs a heatmap display by plotting all the variable pairs in the dataset on the Y-axis and the association measures on the X-axis and encoding the magnitude of association measures with a color.

The `df` argument of the function `pairwise_1d_compare` is used for a dataset. The `measures` argument is for the multiple association measures to be calculated for pairs of variables of `df`. The default measures calculated by the function are Pearson's, Spearman's and Kendall's correlation coefficient, canconical correlation and normalized mutual information. The function also come up with a `var_order` argument which is set to `max_diff` by default and orders the variable pairs on the basis of maximum difference among the magnitude of measures.

Figure \@ref(fig:compare-linear) shows a linear layout comparing multiple association measures for all the variable pairs in the chronic kidney data. The plot shows that the variable pair sc and sod have a magnitude for Pearson's correlation and a magnitude for Spearman's rank correlation suggesting that the relationship among these variables might not be linear and should be looked at in more detail. For variable pairs sc and rbcc, and, sc and pcv, the Spearman's rank correlation has a high magnitude indicating a presence of monotonic relationship for these variable pairs.

```{r compare-linear,fig.width=4, fig.height=6.5, fig.align='center',fig.cap="Comparing multiple association measures using a linear layout. The display has variable pairs on the Y-axis and association measures on the X-axis. The cell corresponding to a variable pair and an association measure has been colored grey showing that the measure is not defined for corresponding pair.",echo=TRUE}


pairwise_1d_compare(df,measures=c("pearson","spearman","kendall","cancor","nmi"))

```


```{r comp_var_pairs, fig.width=6, fig.height=5, fig.align='center', fig.cap = "Scatterplots for numeric variable pairs with high difference among the magnitude of association measures"}


p6 <- ggplot2::ggplot(data = df) + ggplot2::geom_point(ggplot2::aes(y=sc,x=sod))

p7 <- ggplot2::ggplot(data = df) + ggplot2::geom_point(ggplot2::aes(y=sc,x=rbcc))

p8 <- ggplot2::ggplot(data = df) + ggplot2::geom_point(ggplot2::aes(y=sc,x=pcv))

gridExtra::grid.arrange(p6,p7,p8,ncol=3)
```


Figure \@ref(fig:comp_var_pairs) shows scatterplots for the three variable pairs which are picked out from the multiple measures plot. These pairs have a high magnitude difference for Pearson's and Spearman's correlation.

## Conditional Association Plot

The function `pairwise_2d_plot` is used to display a matrix layout of the conditional association for variable pairs in the dataset. The display is produced by splitting the data by a partitioning variable and calculating association for the variable pairs at each level of partitioning variable using `calc_assoc` function with conditioning variable as the `by` argument. The calculated association measures are then displayed using bars in a matrix plot. The height and color of the bars are coded with the value of association measure and the level of the partitioning variable respectively. These displays are efficient for discovering variable pair with high differences among the levels of partitioning variable in the data.

The measures of association calculated for every variable pair at every level of conditioning serve as input to the `pairwise_2d_plot` function. The `lassoc` and `uassoc` arguments look for a tibble of association measures for the lower triangle and the upper triangle of the matrix display respectively. The `uassoc` argument is set to `NULL` by default and uses the same tibble input as used by `lassoc` if not updated. The argument `group_var` is responsible for the grouping variable when plotting the bars. The default value `by` uses the conditioning variable and `measure_type` is useful for displaying bars with height and color of the bars coded by the value of association measure and the type of association measure respectively.

Figure \@ref(fig:cond-assoc) shows a conditional association plot for the chronic kidney data. Each cell corresponding to a variable pair shows two bars which correspond to the association measure (Pearson's correlation for numeric pair, Kendall's tau-b for ordered pair and canonical correlation coefficient for other combination of variables) calculated at the levels of conditioning variable `htn` i.e. whether the patient suffers from hypertension or not.  The dashed line represents the overall association measure. The plot indicates that there is a high difference in the measure (Pearson's correlation) value for the variable pair sc and sod among the patients with and without hypertension. 

```{r cond-assoc,fig.width=6, fig.height=6, fig.align='center', fig.cap="Conditional Association plot for chronic kidney data showing Pearson's correlation for numeric pairs, Kendall's tau-b for ordered pair and canonical correlation for nominal or mixed pairs. The bars in each cell represent the value for asssociation measure colored by the conditioning variable `htn`. The dashed line in each cell represents overall value of the association measure.",echo=TRUE}

cond_assoc <- calc_assoc_by(df, by="htn")
pairwise_2d_plot(cond_assoc)
```

Figure \@ref(fig:high_diff) shows the scatterplot for the variable pair sc and sod spllitted by the conditioning variable htn. The plot indicates that there might be a stronger linear relationship among sc and sod for the people having hypertension compared to patients without hypertension. It is possible to filter out more variable pairs from the conditional association plot having high difference for association measure among the groups and further investigate these pairs.

```{r high_diff, fig.width=6, fig.height=5, fig.align='center', fig.cap = "Scatterplot for variable pair sc and sod splitted by a conditioning variable htn."}

ggplot2::ggplot(data = df) + ggplot2::geom_point(ggplot2::aes(y=sc,x=sod)) + ggplot2::facet_wrap(~htn)
```

<!-- We also provide a functionality for highlighting interesting patterns like Simpson's paradox. Figure \@ref(fig:cond-assoc-sp) shows the matrix plot with highlighted cells for the variable pairs where Simpson's paradox is present. -->

<!-- ```{r cond-assoc-sp,fig.width=5, fig.height=4, fig.align='center',fig.cap="Conditional Association plot with examples of Simpson's paradox"} -->
<!-- pairwise_2d_plot_sp_high(cond_assoc,highlight = "sp") -->
<!-- ``` -->

<!-- The cells can also be highlighted on the basis of a score calculated by the user. This can be done by providing a dataframe with pairs of variables to highlight and a score for highlighting variable pairs. The cells with high score will have a thicker border compared to cells with low score. Figure \@ref(fig:cond-assoc-manual) shows highlighted cells on the basis of a score provided for a subset of variable pairs. -->

<!-- ```{r cond-assoc-manual,fig.width=5, fig.height=4, fig.align='center',fig.cap="Conditional Association plot with manual highlighting"} -->
<!-- highlight <- data.frame(x=c("bill_depth","bill_length","body_mass"), -->
<!--                         y=rep("flipper_length",3),score=c(0.8,0.4,0.1)) -->
<!-- pairwise_2d_plot_sp_high(cond_assoc,highlight = highlight) -->
<!-- ``` -->

We also use linear layouts for displaying conditional association in the package. The function `pairwise_1d_plot` is used for displaying a linear layout of the conditional association for variable pairs in the dataset. The association measures are calculated for every variable pair at each level of partitioning variable using `calc_assoc` function with conditioning variable as the `by` argument. 

The measures are then displayed using a dotplot where color of the dots are coded by the level of the partitioning variable and the variable pairs are ordered by absolute maximum value of association measure for each of the pair of variable. These displays are also efficient for discovering differences among the levels of partitioning variable in the data. With the linear layouts it is easier to omit less relevant pairs of variables by filtering the variables pairs having a higher value for association measures than a threshold.

The measures of association calculated for every variable pair at every level of conditioning serve as input to the `pairwise_1d_plot` function. The `assoc` argument uses a tibble of association measures calculated using `calc_assoc` function with a `by` variable. The argument `group_var` is responsible for the grouping variable when plotting the dots. The default value `by` uses the conditioning variable and `measure_type` is useful for displaying dots with color of the dots coded by the type of association measure. The `var_order` argument is responsible for the ordering of variable pairs in the display. If set to `default` variable pairs are ordered alphabetically and are ordered by absolute maximum value of association measure for every variable pair when set to `max_diff`.

Figure \@ref(fig:linear-cond-assoc) shows a funnel-like linear display for conditional association measures with all the variable pairs on the y-axis, the value of association measure on x-axis and color of the points representing the level of the grouping variable. The linear layout becomes more useful over the matrix layout when the number of variables and number of levels of grouping variable are high.


```{r linear-cond-assoc, fig.width=4, fig.height=5, fig.align='center', fig.cap="Conditional Association plot using linear layout.The display has variable pairs on the Y-axis and the value of association measures on the X-axis. The points corresponding to every variable pair represents the value of association measure for different levels of the conditioning variable and the overall value of association measure.",echo=TRUE}

pairwise_1d_plot(cond_assoc)
```


The `pairwise_2d_plot` function is also useful for comparing various measures using the matrix layout. It plots multiple measures among the variable pairs as bars, where each bar represents one measure of association. Figure \@ref(fig:compare-matrix) shows a matrix layout comparing Pearson's and Spearman's correlation coefficient for the numeric variable pairs in `penguins` data. The plot shows that the value for both the correlation coefficients are very high for `bill_length` and `flipper_length`, `bill_length` and `body_mass`, and `flipper_length` and `body_mass` suggesting a strong linear and montonic relationship among these variable pairs in the dataset.

```{r compare-matrix, fig.width=4, fig.height=4, fig.align='center',fig.cap="Matrix display comparing Pearson's and Spearman's correlation coefficient. All the variable pairs have similar values for both correlations.",echo=TRUE}

df_num <- select(df,where(is.numeric))
pearson <- calc_assoc(df_num)

spearman_assoc <- update_assoc(num_pair = "tbl_cor",
                               num_pair_argList= "spearman",
                               mixed_pair = "tbl_cancor",
                               other_pair = "tbl_nmi")
spearman <- calc_assoc(df_num, types=spearman_assoc)
compare <- rbind(pearson,spearman)
pairwise_2d_plot(compare, group_var = "measure_type")
```
# Section 5: Discussion


We use multiple association measures in a single display for different variable pairs which serves as a comparison tool while exploring association in a dataset and assist in identifying unusual variable pairs. These multiple measures can be displayed in a scatterplot matrix similar to what @tukey1985computer proposed. They suggested that scatterplot matrix of the scagnostics measures, which are measures summarizing a scatterplot, can be used to identify unusual scatterplots or variable pairs. @wilkinson2005graph used this idea with their graph-theoretic scagnostic measures to highlight unusual scatterplots. Similarly, @kuhn2013applied have used this idea in a predictive modeling context. They have produced a scatterplot matrix of the measures between the response and continuous predictors such as Pearson's correlation coefficient, pseudo-$R^2$ from the locally weighted regression model, MIC and Spearman's rank correlation coefficient to explore the predictor importance during feature selection step. These displays show the importance of comparing multiple association measures at once for different variable pairs. 
